Quantitative Auditing of AI Fairness with differentially private Synthetic Data

### 2024-12-11 paper/paper.tex
progress

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Ensuring fairness in AI systems is crucial, but auditing these systems often requires sensitive data, which raises security and privacy concerns. To address this, we introduce a framework that utilizes Differentially Private synthetic data for fairness evaluations. This approach allows auditors to assess AI systems without exposing sensitive information. Through experiments on datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. Our results ... (to be continued)
\end{abstract}

% %%
% %% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% %% Please copy and paste the code instead of the example below.
\begin{CCSXML}
  <ccs2012>
     <concept>
         <concept_id>10002944.10011123.10011124</concept_id>
         <concept_desc>General and reference~Metrics</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10002978.10003029.10011703</concept_id>
         <concept_desc>Security and privacy~Usability in security and privacy</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10002951.10003227.10003351</concept_id>
         <concept_desc>Information systems~Data mining</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10010147.10010257</concept_id>
         <concept_desc>Computing methodologies~Machine learning</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10010147.10010178</concept_id>
         <concept_desc>Computing methodologies~Artificial intelligence</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002978.10002991.10002995</concept_id>
         <concept_desc>Security and privacy~Privacy-preserving protocols</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002951.10003227</concept_id>
         <concept_desc>Information systems~Information systems applications</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002950.10003648.10003688.10003690</concept_id>
         <concept_desc>Mathematics of computing~Contingency table analysis</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
   </ccs2012>
\end{CCSXML}

\ccsdesc[500]{General and reference~Metrics}
\ccsdesc[500]{Security and privacy~Usability in security and privacy}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Computing methodologies~Machine learning}
\ccsdesc[300]{Computing methodologies~Artificial intelligence}
\ccsdesc[300]{Security and privacy~Privacy-preserving protocols}
\ccsdesc[300]{Information systems~Information systems applications}
\ccsdesc[300]{Mathematics of computing~Contingency table analysis}

@@

% TODO: lowercase Diff

@@

% TODO: ai woulld be found to be unfair by 3rd party. example: compas

@@

To conduct audits, auditors may rely on access to real datasets containing sensitive information, raising significant security and privacy concerns. This reliance invites data security attacks on auditors, making them targets of unauthorized access. Analysis on real data runs the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Various technologies have been developed to mitigate these risks, such as Differential Privacy, where the idea is that whether or not any individual is in a dataset has a limited impact.

% TODO: to mitigate use synthetic data. principals of this

@@

We propose a novel auditing framework that leverages Differentially Private synthetic data to evaluate the fairness of AI systems. Our framework lets the auditors assess the fairness of AI systems without exposing sensitive information. Differentially Private synthetic data is a technique that generates synthetic data from real data that preserves some statistical properties of the real data while ensuring Differential Privacy. This technology enables auditing without exposing sensitive information. However, synthetic data also introduces new challenges. Whether or not it accurately preserves the fairness properties of the real data is a critical question.

@@

Our work examines the capabilities of existing Differentially Private synthetic data generation technologies in preserving fairness properties. Through empirical experiments on multiple real datasets, we aim to determine the effectiveness of Differentially Private synthetic data in this respect.

@@

Differential Privacy has risen to become the standard of privacy protection in many data analyses\cite{jiang2021applications}. It offers a formal framework for quantifying privacy guarantees when releasing information derived from sensitive data\cite{dwork2006calibrating,dwork2014algorithmic}. We use Differential Privacy in this work to protect the privacy of individuals in the data in our auditing framework.

Different flavors of Differential Privacy have been proposed over the years\cite{desfontaines2019sok}, such as Gaussian Differential Privacy\cite{dong2022gaussian}, Pufferfish Differential Privacy\cite{kifer2012rigorous}, Bayesian Differential Privacy\cite{triastcyn2020bayesian}, and R\'enyi Differential Privacy\cite{mironov2017renyi}. We consider R\'enyi Differential Privacy specifically in this work.

@@

The generation of synthetic data\cite{raghunathan2021synthetic,lu2023machine} under Differential Privacy constraints allows for data sharing without compromising privacy\cite{tao2021benchmarking}. There have been several techniques developed for Differentially Private synthetic data generation\cite{rosenblatt2020differentially,fan2020relational,bowen2019comparative,bowen2020comparative,arnold2020really,xu2019modeling} in recent years. We use the technique that won the 2018 NIST Differential Privacy Synthetic Data Challenge competition\cite{mckenna2021winning}.

@@

Let $X,Y$ be discrete random variables. Their \emph{mutual information} is $\mi{X}{Y} := \entropy{X} + \entropy{Y} - \entropy{X,Y}$. It quantifies the level of dependence between $X$ and $Y$. By definition\cite{427884}, the joint entropy is greater than or equal to the marginal entropies $\entropy{X,Y} \geq \text{max}(\entropy{X},\entropy{Y})$. Hence, we have an upper bound of mutual information $\mi{X}{Y} \leq \text{min}(\entropy{X}, \entropy{Y})$.

@@

For fairness measures\cite{yuan2024ensuring,pessach2022review}, let $Y$ denote the ground truth of an outcome, let $\hat{Y}$ denote the predicated result of an outcome, let $S$ denote the protected attribute, and let $\epsilon$ denote some threshold. $Y, \hat{Y}, S$ are binary. For non-binary prediction, such as a score, we use $\hat{V}$.

@@

% TODO: remove unconsidered

% TODO: rename equalized odds 1,2

\begin{table*}[h]
\caption{Fairness measures.}
\label{tab:measures}
\begin{tabular}{llc}
\toprule
\textbf{Category} & \textbf{Fairness Measure} & \textbf{Definition} \\
\midrule
\multirow{1}{*}{Independence}
% & Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
& Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
% & Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
% & Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\multirow{2}{*}{Separation}
& \multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
& & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\multirow{2}{*}{Sufficiency}
& \multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
& & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
% & Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
% & Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\multirow{1}{*}{N/A}
& Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
% & Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
% & Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{R\'enyi Differential Privacy}

% \begin{definition}[Differential Privacy (DP) \cite{dwork2006calibrating,dwork2014algorithmic,mckenna2021winning}]\label{def:rdp}
% A randomized mechanism $M$ satisfies $(\epsilon,\delta)$-DP if, for all databases $\db_1,\db_2$ that differ in exactly one row and for all subsets $S$ of $R$, we have
% \[
% \prob{M(\db_1) \in S} \leq e^\epsilon \prob{M(\db_2) \in S} + \delta
% \]
% \end{definition}

@@

Let $\db_1,\db_2$ be two databases. They are \emph{neighbors}, denoted $\db_1 \sim \db_2$, if $\abs{\mathcal{R}_{\db_1} \Delta \mathcal{R}_{\db_2}} \in \set{1,2}$ where $\Delta$ denotes symmetric difference; that is, either one database contains an extra row, or both databases have all but one row in common. In other words, they differ in exactly one row.

@@

The task of Differentially Private synthetic data\cite{Ullman2022,McKenna2022,NIST_Differentially_Private_Synthetic_Data,YouTube_Differentially_Private_Synthetic_Data} is, given a database $\db$, adding some noise to marginals of $\db$ such that it satisfies some Differential Privacy guarantees and outputting another database $\db'$, such that the $L^1$-norm errors between marginals of $\db$ and $\db'$ is small; that is, their marginals $\mu_{\db},\mu_{\db'}$ are similar.

@@

\begin{table}[h]
\caption{Example marginals.}
\label{tab:marginal-example}
\centering
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Marginal of original data.}
\begin{tabular}{cc}
\toprule
Attributes & Count \\
\midrule
Male,White & 24 \\
Female,White & 33 \\
Male,Black & 13 \\
Female,Black & 47 \\
\bottomrule
\end{tabular}
\end{subtable}
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Marginal of synthetic data.}
\begin{tabular}{cc}
\toprule
Attributes & Count \\
\midrule
Male,White & 22 \\
Female,White & 35 \\
Male,Black & 10 \\
Female,Black & 46 \\
\bottomrule
\end{tabular}
\end{subtable}
\end{table}

@@

% TODO: add a overarching example for methodologies

We employed the tools of the winner of the 2018 NIST Differential Privacy Synthetic Data Challenge competition\cite{NIST2018} by \cite{nist_ryan_2018,McKenna_Private-PGM_2021,mckenna2021winning,mckenna2019graphical} and the fairness checker tool from our previous research\cite{yuan2024ensuring}.

@@

@@

The synthesis framework is three-fold, namely, select-measure-generate\cite{McKenna2022,YouTube_Differentially_Private_Synthetic_Data}. We first select the important marginals to preserve, measure them by adding Differential Privacy noise, and then generate synthetic data.

@@

To circumvent this complexity explosion, instead, \cite{mckenna2021winning} devised a technique where the mutual information of all the database attribute pairs is calculated, and then a maximum spanning tree is identified with edge weights being the mutual information to obtain a skeleton tree-shaped Markov random field.

For the competition, \cite{mckenna2021winning} further manually added certain cliques based on his investigation of the competition dataset. For example, they manually added the clique (sex,city,income). In addition, they would add some edges based on some sophisticated heuristics tailored to that particular dataset. Meanwhile, his other apporach where the auditor does access the real dataset does not fit our framework.

@@

% TODO:
% for synth to good : depend on domain knowledge, tractibility
% synth algos use dependencies
% we DO NOT use this. we do vanilla
% experiment didnt developed heuristics:
% 1. lack of domain knowledge
% 2. different prediction target. cannot tailor a priori

For the measure step, we followed the examples provided in the tool's repository. Gaussian noises are added to the selected marginals. Half of the privacy budget is spent on all 1-way marginals and the other half on the selected cliques. These marginals are then fed to the tool to fit the Markov random field. By \cite{mckenna2021winning}, this procedure satisfies $(\alpha,\frac{\alpha}{2 \sigma^2})$-RDP for all $\alpha \geq 1$.

@@

For example, if the sensitive attribute is sex and the protected group is female, the protected group predicate would be $S := r(sex) = Female$. This can be easily implemented in Python as a comparison function.

@@

% TODO: move this to their respective subsection

% TODO: reiterate methodologies

We looked at several publicly available datasets, such as Adult\cite{adult_2,Kaggle_Adult_Census_Income}, COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset}, and Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction}. The Adult dataset comes from the 1994 census in the United States and contains about 30000 individuals. The COMPAS dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The Diabetes dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. Since these datasets all have binary outcomes, we did not consider the case of non-binary outcomes in the experiments.

The fairness checker evaluates datasets based on multiple fairness metrics, such as demographic parity and equalized odds. These metrics are computed on some sensitive attributes, predicted outcomes, and ground truths. Examples of sensitive attributes are race and sex. Examples of predicted outcomes and ground truths are loan approval and criminal recidivism.

By comparing these measures between the synthetic and original datasets, we aim to ensure that the synthetic data preserves the fairness properties of the original data. The comparison process is three-fold. It goes as follows.

The dataset is first processed so it can be fed into the synthetic data generator. Some marginals are selected as described in the Section~\ref{sec:method}, and the synthetic data generator model is fitted to the original data according to the marginals. Then the generator is run multiple times to obtain multiple sets of synthetic data.

@@

Several AI models and both the original dataset and the rounds of synthetic datasets are fed to the fairness checker. Sensitive attributes are identified based on manual examination with common sense or by referring to \cite{pessach2022review}. Then, all applicable fairness measures are computed using the checker for both the original and the synthetic.

Finally, we analyze the discrepancies between the fairness properties of the original and the synthetic by calculating the difference of their perspective fairness measure values. The average of the differences serves as a summary of the analysis.

@@

% TODO: dont mention shape of graph

For the Adult income dataset, the shape of the maximum spanning tree is very shallow, almost resembling a star; it has one internal node and all but one of the leaves have a depth of one. After introducing edges according to our heuristic, we observed an increase in the pairwise edges of the leaves, forming many 3-cliques and two 4-cliques. The resulting graph is shown in Figure~\ref{fig:adult_mst}.

Marginals based on this graph are then passed to the synthetic data generator for model fitting.

After generating ten rounds of synthetic data and passing them to the checker, their fairness measure values are averaged. Then, we compare them against the values of the original data. The results are shown in Table~\ref{tab:adult_score}.

@@

@@

% Marginals of this graph are then too passed to the synthetic data generator for fitting.

% We ran the same workflow as adult dataset for the COMPAS dataset. The comparison results are shown in Table~\ref{tab:compas_score}.

@@

@@

% The same process is conducted to fit the synthetic data generator model.

% The same workflow was done as on previous datasets. The comparison results are shown in Table~\ref{tab:diabetes_score}.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Quantitative Auditing of AI Fairness with differentially private Synthetic Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Ensuring fairness in AI systems is crucial, but auditing these systems often requires sensitive data, which raises security and privacy concerns. To address this, we introduce a framework that utilizes differentially private synthetic data for fairness evaluations. This approach allows auditors to assess AI systems without exposing sensitive information. Through experiments on datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. Our results ... (to be continued)
\end{abstract}

% %%
% %% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% %% Please copy and paste the code instead of the example below.
\begin{CCSXML}
  <ccs2012>
     <concept>
         <concept_id>10002944.10011123.10011124</concept_id>
         <concept_desc>General and reference~Metrics</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10002978.10003029.10011703</concept_id>
         <concept_desc>Security and privacy~Usability in security and privacy</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10002951.10003227.10003351</concept_id>
         <concept_desc>Information systems~Data mining</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10010147.10010257</concept_id>
         <concept_desc>Computing methodologies~Machine learning</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10010147.10010178</concept_id>
         <concept_desc>Computing methodologies~Artificial intelligence</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002978.10002991.10002995</concept_id>
         <concept_desc>Security and privacy~privacy-preserving protocols</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002951.10003227</concept_id>
         <concept_desc>Information systems~Information systems applications</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10002950.10003648.10003688.10003690</concept_id>
         <concept_desc>Mathematics of computing~Contingency table analysis</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
   </ccs2012>
\end{CCSXML}

\ccsdesc[500]{General and reference~Metrics}
\ccsdesc[500]{Security and privacy~Usability in security and privacy}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Computing methodologies~Machine learning}
\ccsdesc[300]{Computing methodologies~Artificial intelligence}
\ccsdesc[300]{Security and privacy~privacy-preserving protocols}
\ccsdesc[300]{Information systems~Information systems applications}
\ccsdesc[300]{Mathematics of computing~Contingency table analysis}

@@

@@

Auditing can evaluate the fairness of AI systems, sometimes revealing cases of unfairness in them. For example, investigative journalists at ProPublica found that the COMPAS system, which is used to predict recidivism in the criminal justice system in the United States, is biased against certain demographic groups. This highlights the need for fairness auditing conducted by third parties.

@@

To conduct audits, auditors may rely on access to real datasets containing sensitive information, raising significant security and privacy concerns. This reliance invites data security attacks on auditors, making them targets of unauthorized access. Analysis on real data runs the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Various technologies have been developed to mitigate these risks, such as differential privacy, where the idea is that whether or not any individual is in a dataset has a limited impact.

To mitigate these risks, the use of synthetic data has been proposed. Synthetic data are fake data that is artificially generated from real data. The principle of this technology is to preserve some statistical properties of the real data while protecting the privacy of the individuals in the real data. Among the generation techniques, differentially private synthetic data offers a way to investigate the data properties under differential privacy guarantees.

@@

We propose a novel auditing framework that leverages differentially private synthetic data to evaluate the fairness of AI systems. Our framework lets the auditors assess the fairness of AI systems without exposing sensitive information. Differentially private synthetic data is a technique that generates synthetic data from real data that preserves some statistical properties of the real data while ensuring differential privacy. This technology enables auditing without exposing sensitive information. However, synthetic data also introduces new challenges. Whether or not it accurately preserves the fairness properties of the real data is a critical question.

@@

Our work examines the capabilities of existing differentially private synthetic data generation technologies in preserving fairness properties. Through empirical experiments on multiple real datasets, we aim to determine the effectiveness of differentially private synthetic data in this respect.

@@

Differential privacy has risen to become the standard of privacy protection in many data analyses\cite{jiang2021applications}. It offers a formal framework for quantifying privacy guarantees when releasing information derived from sensitive data\cite{dwork2006calibrating,dwork2014algorithmic}. We use differential privacy in this work to protect the privacy of individuals in the data in our auditing framework.

Different flavors of differential privacy have been proposed over the years\cite{desfontaines2019sok}, such as Gaussian differential privacy\cite{dong2022gaussian}, Pufferfish differential privacy\cite{kifer2012rigorous}, Bayesian differential privacy\cite{triastcyn2020bayesian}, and R\'enyi differential privacy\cite{mironov2017renyi}. We consider R\'enyi differential privacy specifically in this work.

@@

The generation of synthetic data\cite{raghunathan2021synthetic,lu2023machine} under differential privacy constraints allows for data sharing without compromising privacy\cite{tao2021benchmarking}. There have been several techniques developed for differentially private synthetic data generation\cite{rosenblatt2020differentially,fan2020relational,bowen2019comparative,bowen2020comparative,arnold2020really,xu2019modeling} in recent years. We use the technique that won the 2018 NIST differential privacy Synthetic Data Challenge competition\cite{mckenna2021winning}.

@@

Let $X,Y$ be discrete random variables. Their \emph{mutual information} is $\mi{X}{Y} := \entropy{X} + \entropy{Y} - \entropy{X,Y}$. It quantifies the level of dependence between $X$ and $Y$.
% By definition\cite{427884}, the joint entropy is greater than or equal to the marginal entropies $\entropy{X,Y} \geq \text{max}(\entropy{X},\entropy{Y})$. Hence, we have an upper bound of mutual information $\mi{X}{Y} \leq \text{min}(\entropy{X}, \entropy{Y})$.

@@

For fairness measures\cite{yuan2024ensuring,pessach2022review}, let $Y$ denote the ground truth of an outcome, let $\hat{Y}$ denote the predicated result of an outcome, let $S$ denote the protected attribute, and let $\epsilon$ denote some threshold. $Y, \hat{Y}, S$ are binary.

@@

\begin{table*}[h]
\caption{Fairness measures.}
\label{tab:measures}
\begin{tabular}{lll}
\toprule
\textbf{Category} & \textbf{Fairness Measure} & \textbf{Definition} \\
\midrule
\multirow{1}{*}{Independence}
% & Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
& Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
% & Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
% & Mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\multirow{2}{*}{Separation}
& \multirow{1}{*}{Equalized Odds (False Positive)} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
& \multirow{1}{*}{Equalized Odds (True Positive)} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\multirow{2}{*}{Sufficiency}
& \multirow{1}{*}{Conditional Use Accuracy Equality (True Positive)} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
& \multirow{1}{*}{Conditional Use Accuracy Equality (True Negative)} & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
% & Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
% & Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\multirow{1}{*}{N/A}
& Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
% & Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
% & Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{R\'enyi differential privacy}

% \begin{definition}[differential privacy (DP) \cite{dwork2006calibrating,dwork2014algorithmic,mckenna2021winning}]\label{def:rdp}
% A randomized mechanism $M$ satisfies $(\epsilon,\delta)$-DP if, for all databases $\db_1,\db_2$ that differ in exactly one row and for all subsets $S$ of $R$, we have
% \[
% \prob{M(\db_1) \in S} \leq e^\epsilon \prob{M(\db_2) \in S} + \delta
% \]
% \end{definition}

@@

Let $\db_1,\db_2$ be two databases. They are \emph{neighbors}, denoted $\db_1 \sim \db_2$, if $\abs{\mathcal{R}_{\db_1} \Delta \mathcal{R}_{\db_2}} = 1$ where $\Delta$ denotes symmetric difference; that is, one database contains an extra row than the other. In other words, they differ in exactly one row.

@@

The task of differentially private synthetic data\cite{Ullman2022,McKenna2022,NIST_Differentially_Private_Synthetic_Data,YouTube_Differentially_Private_Synthetic_Data} is, given a database $\db$, adding some noise to marginals of $\db$ such that it satisfies some differential privacy guarantees and outputting another database $\db'$, such that the $L^1$-norm errors between marginals of $\db$ and $\db'$ is small; that is, their marginals $\mu_{\db},\mu_{\db'}$ are similar.

@@

\begin{table}[h]
\caption{Example marginals.}
\label{tab:marginal-example}
\centering
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Marginal of original data.}
\begin{tabular}{lr}
\toprule
Attributes & Count \\
\midrule
Male,White & 24 \\
Female,White & 33 \\
Male,Black & 13 \\
Female,Black & 47 \\
\bottomrule
\end{tabular}
\end{subtable}
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Marginal of synthetic data.}
\begin{tabular}{lr}
\toprule
Attributes & Count \\
\midrule
Male,White & 22 \\
Female,White & 35 \\
Male,Black & 10 \\
Female,Black & 46 \\
\bottomrule
\end{tabular}
\end{subtable}
\end{table}

@@

We employed the tools of the winner of the 2018 NIST differential privacy Synthetic Data Challenge competition\cite{NIST2018} by \cite{nist_ryan_2018,McKenna_private-PGM_2021,mckenna2021winning,mckenna2019graphical} and the fairness checker tool from our previous research\cite{yuan2024ensuring}.

@@

We shall illustrate the workflow of our methodology with an example of the COMPAS dataset, which contains attributes such as sex and race with the prediction target attribute being recidivism.

@@

The synthesis framework is three-fold, namely, select-measure-generate\cite{McKenna2022,YouTube_Differentially_Private_Synthetic_Data}. We first select the important marginals to preserve, measure them by adding differential privacy noise, and then generate synthetic data.

@@

To circumvent this complexity explosion, instead, \cite{mckenna2021winning} devised a technique where the mutual information of all the database attribute pairs is calculated, and then a maximum spanning tree is identified with edge weights being the mutual information to obtain a skeleton tree-shaped Markov random field. The marginals are then selected based on the edges of this tree.

In the COMPAS example, we may select the marginal (sex,race).

For the competition, \cite{mckenna2021winning} further manually added certain cliques based on his investigation of the competition dataset. For example, they manually added the clique (sex,city,income). In addition, they would add some edges based on some sophisticated heuristics tailored to that particular dataset.

We do not consider these additional heuristics for the generality of our approach. Meanwhile, his other apporach where the auditor does access the real dataset does not fit our framework.

While synthetic data generation methods often leverage domain knowledge to improve accuracy, we chose to employ a vanilla approach with only the maximum spanning tree method without incorporating additional heuristics. This decision was made due to two main factors: first, the need for more specific domain knowledge about the dataset makes it difficult to identify meaningful dependencies; second, prediction targets may differ from dataset to dataset, which means that tailored heuristics could not be effectively generalized. By avoiding dataset-specific adjustments, our method remains general and applicable to a wide range of datasets. Furthermore, the bare tree structure makes the method more tractable and computationally efficient.

@@

For the measure step, we followed the examples provided in the tool's repository. Gaussian noises are added to the selected marginals. Half of the privacy budget $\epsilon = 1$ is spent on all 1-way marginals and the other half on the selected cliques. These marginals are then fed to the tool to fit the Markov random field. By \cite{mckenna2021winning}, this procedure satisfies $(\alpha,\frac{\alpha}{2 \sigma^2})$-RDP for all $\alpha \geq 1$.

In the COMPAS example, the selected the marginals are perturbed by Gaussian noise, as shown in Table~\ref{tab:marginal-perturbed}.

\begin{table}[h]
\caption{Example marginals.}
\label{tab:marginal-perturbed}
\centering
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Original marginal.}
\begin{tabular}{lr}
\toprule
Attributes & Count \\
\midrule
Female,African-American & 1537 \\
Female,Asian & 8 \\
Female,Caucasian & 1465 \\
Female,Hispanic & 215 \\
Female,Native American & 15 \\
Female,Other & 143 \\
Male,African-American & 8254 \\
Male,Asian & 63 \\
Male,Caucasian & 4621 \\
Male,Hispanic & 1236 \\
Male,Native American & 42 \\
Male,Other & 717 \\
\bottomrule
\end{tabular}
\end{subtable}
\begin{subtable}[t]{0.45\linewidth}
\centering
\caption{Perturbed marginal.}
\begin{tabular}{lr}
\toprule
Attributes & Count \\
\midrule
Female,African-American & 1545.7885 \\
Female,Asian & 4.3236 \\
Female,Caucasian & 1444.7303 \\
Female,Hispanic & 208.2250 \\
Female,Native American & 26.7479 \\
Female,Other & 150.8175 \\
Male,African-American & 8296.6274 \\
Male,Asian & 92.1107 \\
Male,Caucasian & 4643.0108 \\
Male,Hispanic & 1274.0831 \\
Male,Native American & 28.9542 \\
Male,Other & 710.3473 \\
\bottomrule
\end{tabular}
\end{subtable}
\end{table}

@@

In the COMPAS example, if the sensitive attribute is sex and the protected group is female, the protected group predicate would be $S := r(sex) = Female$. This can be easily implemented in Python as a comparison function.

@@

We looked at several publicly available datasets, such as Adult\cite{adult_2,Kaggle_Adult_Census_Income}, COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset}, and Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction}. These datasets are known to have fairness issues\cite{pessach2022review}.

The fairness checker evaluates datasets based on multiple fairness metrics, such as demographic parity and equalized odds. By comparing these measures between the synthetic and original datasets, we aim to ensure that the synthetic data preserves the fairness properties of the original data. The comparison process is three-fold. It goes as follows.

The dataset is first processed so it can be fed into the synthetic data generator. Some marginals are selected according to their mutual information, as described in the Section~\ref{sec:method}, and the synthetic data generator model is fitted to the original data according to the selected marginals. Then the generator is run 100 times to obtain 100 sets of synthetic data.

@@

The AI models and both the original dataset and synthetic datasets are then fed to the fairness checker. Sensitive attributes are identified based on manual examination with common sense or by referring to \cite{pessach2022review}. Then, applicable fairness measures are computed using the checker for both the original and the synthetic.

@@

The Adult\cite{adult_2,Kaggle_Adult_Census_Income} dataset comes from the 1994 census in the United States and contains about 30000 individuals. The dataset has 14 attributes, such as sex, age, and education. The task is to predict whether an individual earns more than \$50,000 a year. We set the sensitive attribute as sex and the protected group as female. The results are shown in Table~\ref{tab:adult_score}.

% For the Adult income dataset, the shape of the maximum spanning tree is very shallow, almost resembling a star; it has one internal node and all but one of the leaves have a depth of one. After introducing edges according to our heuristic, we observed an increase in the pairwise edges of the leaves, forming many 3-cliques and two 4-cliques. The resulting graph is shown in Figure~\ref{fig:adult_mst}.

@@

The COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset} dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The dataset has 10 applicable attributes, such sex, age, and priors count. The task is to predict whether an individual will re-offend. We set the sensitive attribute as sex and the protected group as female. The results are shown in Table~\ref{tab:compas_score}.

@@

@@

The Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction} dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. The dataset has 8 attributes, such as gender, age, and smoking history. The task is to predict whether an individual will be readmitted. We set the sensitive attribute as gender and the protected group as female. The results are shown in Table~\ref{tab:diabetes_score}.

@@
